---
title: "Heatmap DE genes in IL21"
author: "Dan Griffiths"
date: "2023-11-14"
output:
  html_document: default
---

### Load libraries

```{r libraries, message = FALSE}
library(limma)
library(Glimma)
library(tidyverse)
library(DESeq2)
library(org.Hs.eg.db)
library(pheatmap)
library(EnhancedVolcano)
library(ComplexHeatmap)
library(gplots)
library(msigdbr)
```

```{r}
# rds <- load("~/Downloads/subread/FeatCounts_Rsub.RData")
# counts <- as.matrix(counts$counts)
# # Find column indices that don't contain 'markdup' in the names
# selected_columns <- grep("markdup", colnames(counts), invert = TRUE)
# 
# # Select the columns based on the indices
# filtered_matrix <- counts[, selected_columns]
# 
# # Remove ".sorted.bam" from column names
# new_colnames <- sub("\\.sorted\\.bam", "", colnames(filtered_matrix))
# 
# # Assign the modified column names back to the matrix
# colnames(filtered_matrix) <- new_colnames
# 
# counts <- filtered_matrix
```


### Import raw counts, coldata
```{r, message=FALSE}
set.seed(42)
coldata <- readr::read_csv('./source_data/coldata.csv') %>% as.data.frame()
identical(colnames(counts), coldata$sample)
```


```{r}
counts <- readr::read_rds('source_data/DDS_Full_counts.RDS') %>% counts(.) %>% as.matrix()
rownames(counts) <- mapIds(org.Hs.eg.db, keys = rownames(counts), column = "ENTREZID", keytype = "ENSEMBL")
# Create gene list to map from ensembl ID to 'symbol' ID -- more readable format
gene.list <- data.frame(entrez.id = rownames(counts))
gene.list <- gene.list %>%
  dplyr::mutate(
    SYMBOL = mapIds(org.Hs.eg.db, keys = gene.list$entrez.id, column = "SYMBOL", keytype = "ENTREZID"),
    GENETYPE = mapIds(org.Hs.eg.db, keys = gene.list$entrez.id, column = "GENETYPE", keytype = "ENTREZID")
  )

# db.gs.c7 <- msigdbr(species = "human", category = "C7", subcategory = "IMMUNESIGDB")
# # unique(db.gs.c7$gs_subcat)
# 
# 
# # Genes removed if protein coding,
# filtered.gene.list <- gene.list %>%
#   dplyr::filter(
#     !is.na(SYMBOL),
#     !duplicated(SYMBOL),
#     GENETYPE == "protein-coding"
#   )

# filtered.gene.list <- filtered.gene.list %>%
#   filter(entrez.id %in% unique(db.gs.c7$human_entrez_gene))

# Rename gene IDs from ensembl to symbol
counts <- counts[row.names(counts) %in% filtered.gene.list$entrez.id, ]
rownames(counts) <- filtered.gene.list$SYMBOL[match(rownames(counts), filtered.gene.list$entrez.id)]

```




### Prepare and filter counts
```{r, message=FALSE}
# # Create gene list to map from ensembl ID to 'symbol' ID -- more readable format
# gene.list <- data.frame(entrez.id = rownames(counts))
# gene.list <- gene.list %>%
#   dplyr::mutate(
#     SYMBOL = mapIds(org.Hs.eg.db, keys = gene.list$entrez.id, column = "SYMBOL", keytype = "ENTREZID"),
#     GENETYPE = mapIds(org.Hs.eg.db, keys = gene.list$entrez.id, column = "GENETYPE", keytype = "ENTREZID")
#   )
# 
# db.gs.c7 <- msigdbr(species = "human", category = "C7", subcategory = "IMMUNESIGDB")
# # unique(db.gs.c7$gs_subcat)
# 
# 
# # Genes removed if protein coding,
# filtered.gene.list <- gene.list %>%
#   dplyr::filter(
#     !is.na(SYMBOL),
#     !duplicated(SYMBOL),
#     GENETYPE == "protein-coding"
#   )
# 
# # filtered.gene.list <- filtered.gene.list %>%
# #   filter(entrez.id %in% unique(db.gs.c7$human_entrez_gene))
# 
# # Rename gene IDs from ensembl to symbol
# counts <- counts[row.names(counts) %in% filtered.gene.list$entrez.id, ]
# rownames(counts) <- filtered.gene.list$SYMBOL[match(rownames(counts), filtered.gene.list$entrez.id)]
```

### Generate DESeq2 object
IL7 removed from analysis completely.
```{r, message=FALSE, warning=FALSE}
dds <- DESeqDataSetFromMatrix(counts,coldata, design = ~condition)
#dds <- dds[ , !dds$condition %in% c("IL7","UST") ] 
dds$condition <- droplevels(dds$condition)
smallestGroupSize <- 3
keep <- rowSums(counts(dds) >= 5) >= smallestGroupSize
dds <- dds[keep,]
dds <- DESeq(dds)
vst <- varianceStabilizingTransformation(dds)
vsn::meanSdPlot(assay(vst))
```

```{r,fig.height=20,fig.width=10}
vst_data <- assay(vst)
topVarGenes <- order(rowVars(vst_data), decreasing = TRUE)
mat  <- vst_data[1:50, ]
mat  <- mat - rowMeans(mat)

anno <- as.data.frame(colData(dds))
annocols1 <- paste(c(unique(anno$condition)))
annocols2 <- paste(c(unique(anno$healthy_donor)))
vip.cols1 <- RColorBrewer::brewer.pal(name = "Dark2", n = length(annocols1))
vip.cols2 <- RColorBrewer::brewer.pal(name = "Set2", n = length(annocols2))
names(vip.cols1) = sort(unique(anno$condition))
names(vip.cols2) = sort(unique(anno$healthy_donor))

HA <- columnAnnotation(df = anno %>% dplyr::select(c("condition","healthy_donor")), col = list(condition=vip.cols1,healthy_donor=vip.cols2))

Heatmap(mat, 
        name = "Relative variance",
        column_title = "Common genes differentially expressed in IL21 vs IL15, and IL21 vs IL2; \n variance stabilised expression heatmap \n gene expression relative to the mean", 
        top_annotation = HA, 
        cluster_columns = FALSE, 
        column_split = anno$healthy_donor,
        cluster_rows = TRUE
        )

```


### DESeq2 results pairwise each condition
```{r}
res.il15 <- results(dds,contrast = c("condition","IL21","IL15"))
res.sig.il15 <- subset(res.il15, padj < 0.05)
lfc.il15 <- res.sig.il15[ order(res.sig.il15$log2FoldChange, decreasing=TRUE), ]
summary(lfc.il15)
lfc.il15
res.il2 <- results(dds,contrast = c("condition","IL21","IL2"))
res.sig.il2 <- subset(res.il2, padj < 0.05)
lfc.il2 <- res.sig.il2[ order(res.sig.il2$log2FoldChange, decreasing=TRUE), ]
summary(lfc.il2)
lfc.il2


lfc.il15.df <- lfc.il15 %>% as.data.frame() %>% rownames_to_column(.,var = 'GENE_ID')
lfc.il2.df <- lfc.il2 %>% as.data.frame() %>% rownames_to_column(.,var = 'GENE_ID')
write_csv(lfc.il15.df, "lfc_il15_results.csv")
write_csv(lfc.il2.df, "lfc_il2_results.csv")
```

### Venn diagram common genes
```{r}
## Data preparation
il15.degs <- row.names(lfc.il15)
il2.degs <- row.names(lfc.il2)
list.degs <- paste(c(il15.degs,il2.degs))
list.degs <- unique(list.degs)

# Create a Venn-diagram given just the list of gene-names for both sets
venn(list("IL15vsIL21" = il15.degs, "IL2vsIL21" = il2.degs), )
common_genes <- Reduce(intersect, list(il2.degs, il15.degs))
```

```{r,fig.height=15,fig.width=10}
vst_data <- assay(vst)
topVarGenes <- order(rowVars(vst_data), decreasing = TRUE)
top_20 <- common_genes[1:10]
bottom_20 <- tail(common_genes, n=10)
short.common <- paste(c(top_20,bottom_20))
mat  <- vst_data[short.common, ]
mat  <- mat - rowMeans(mat)

anno <- as.data.frame(colData(dds))
annocols1 <- paste(c(unique(anno$condition)))
annocols2 <- paste(c(unique(anno$healthy_donor)))
vip.cols1 <- RColorBrewer::brewer.pal(name = "Dark2", n = length(annocols1))
vip.cols2 <- RColorBrewer::brewer.pal(name = "Set2", n = length(annocols2))
names(vip.cols1) = sort(unique(anno$condition))
names(vip.cols2) = sort(unique(anno$healthy_donor))

HA <- columnAnnotation(df = anno %>% dplyr::select(c("condition","healthy_donor")), col = list(condition=vip.cols1,healthy_donor=vip.cols2))

Heatmap(mat, 
        name = "Relative variance",
        column_title = "Common genes differentially expressed in IL21 vs IL15, and IL21 vs IL2; \n variance stabilised expression heatmap \n gene expression relative to the mean", 
        top_annotation = HA, 
        cluster_columns = FALSE, 
        column_split = anno$healthy_donor,
        cluster_rows = TRUE
        )

```


FCRL6 -->
" In contrast to FCRL1–5, FCRL6 is distinctly expressed outside the B lineage by cytotoxic T and NK lymphocytes"
https://www.frontiersin.org/articles/10.3389/fimmu.2020.575175/full


https://www.tandfonline.com/doi/abs/10.1080/14728222.2018.1472768



TIMP-1 -->
"In addition to its inhibitory role against most of the known MMPs, the encoded protein is able to promote cell proliferation in a wide range of cell types, and may also have an anti-apoptotic function. Transcription of this gene is highly inducible in response to many cytokines and hormones." https://www.genecards.org/cgi-bin/carddisp.pl?gene=TIMP1&keywords=timp1



```{r}
# plot_data <- data.frame(Category = c("IL21vsIL2", "IL21vsIL15", "Common"),
#                         Count = c(length(IL21vsIL2_genes), length(IL21vsIL15_genes), length(common_genes)))
# 
# # Create a bar plot
# ggplot(plot_data, aes(x = Category, y = Count, fill = Category)) +
#   geom_bar(stat = "identity") +
#   labs(title = "Common Genes between IL21 vs IL2 and IL21 vs IL15",
#        x = "Comparison",
#        y = "Number of Genes") +
#   theme_minimal()
```



```{r,fig.height=10,fig.width=15}
EnhancedVolcano(res.il2,
    lab = rownames(res.il2),
    x = 'log2FoldChange',
    y = 'pvalue',
    title = 'IL21 versus IL2',
    pCutoff = 5e-4,
    FCcutoff = 2,
    pointSize = 3.0,
    labSize = 6.0)

EnhancedVolcano(res.il15,
    lab = rownames(res.il15),
    x = 'log2FoldChange',
    y = 'pvalue',
    title = 'IL21 versus IL15',
    pCutoff = 5e-4,
    FCcutoff = 2,
    pointSize = 3.0,
    labSize = 6.0)
```


## R session info

```{r}
sessionInfo()
```
